<!-- LTeX: language=en-US -->
```{r setup, include=FALSE, cache=FALSE}
source("_common.R")
```

# Organization of the Course {-}

### Timetable {-}

```{r, echo=FALSE}
suppressPackageStartupMessages(library("kableExtra"))
suppressPackageStartupMessages(library("tidyverse"))

timetable <- tibble("Day"          = c("Thu.",       "Thu."),
                    "Time"         = c("8:30-10:00", "12:00-13:30"),
                    "Lecture Hall" = c("NB 301",         "GH"))

timetable %>% kbl() %>%  kable_styling()
```


### Lecture Material  

* This online script
* [eWhiteboard](https://www.dropbox.com/scl/fi/cekkrg29r1toqha35wdek/eWhiteboard_2627_annotated.pdf?rlkey=ddrcbxw5y4nlp59j8p5edvlgl&st=jaxomtsl&dl=0)

### Literature {-}

The script is self-contained. To prepare well for the exam, it's a good idea to work with this script. 

Further textbooks I can recommend:  

* *All of Statistics: A Concise Course in Statistical Inference*, by Larry Wasserman
* *Econometrics*, by F. Hayashi
* *Econometrics*, by Bruce E. Hansen
* *Probability and Statistics for Economists*, by Bruce E. Hansen
* *Econometric theory and methods*, by R. Davidson and J.G. MacKinnon
<!-- * *Introduction to econometrics*, by J. Stock and M.W. Watson
  * TODO — KU: the e-book link here was a Bonn BONNUS/ULB catalogue record.
    Replace with the KU library equivalent if this entry is restored. -->
<!-- * *A primer in econometric theory*, by J. Stachurski -->


### Information on the Exam {-}

* Written exam (60-90 minutes)
<!-- 
* **10** multiple-choice problems and
   * **2** free-text problems (each with sub-questions)
* In the exam you are allowed to use
   * a *handwritten* cheat-sheet (**one side** of a DIN A4 page).
   * a non-programmable scientific calculator.
* Please, do **not** use a pencil.
* Exam dates and times are published by the examinations office (TODO — KU: add link) 
-->
<!-- * There will be a **Mock Exam** in the last lecture weeks -->

<!-- TODO — KU: the mock exam below linked to the Bonn WS 25/26 Dropbox
     (Mock_Exam1.pdf). Restore with a KU mock exam when one exists.

* **Mock Exam** with 5 multiple-choice problems and 1 free-text problem: [Download](ADD-KU-LINK)
   * Note: The actual exam will contain **10** multiple-choice problems and **2** free-text problems.
-->



