<!-- LTeX: language=en-US -->
```{r setup, include=FALSE, cache=FALSE}
source("_common.R")
```

# Probability {#ReviewStats}

The content of this chapter follows closely (often verbatim) the fantastic book by @Wasserman2004. 

## Basics in Probability Theory

Probability is the mathematical language for quantifying uncertainty. We can apply probability theory to a diverse set of problems, from coin flipping to the analysis of econometric problems. The starting point is to specify the **sample space**, that is, the set of possible outcomes.


### Sample Spaces and (Elementary) Events
The **sample space** $\Omega,$ is the set of possible outcomes of an experiment. Points $\omega$ in $\Omega$ are called **sample outcomes** or **realizations** or **elementary events**. **Events** are subsets of $\Omega$. 



**Example:** If we toss a coin twice then $\Omega=\{H H, H T, T H, T T\}.$ The event that the first toss is heads is $A=\{H H, H T\}.$


**Example:** Let $\omega$ be the outcome of a measurement of some physical quantity, for example, temperature. Then $\Omega=\mathbb{R}=(-\infty, \infty).$ The event that the measurement is larger than 10 but less than or equal to 23 is $A=(10,23].$


**Example:**  If we toss a coin forever then the sample space is the infinite set $\Omega=\left\{\omega=\left(\omega_{1}, \omega_{2}, \omega_{3}, \ldots,\right)|\omega_{i} \in\{H, T\}\right\}$ Let $A$ be the event that the first head appears on the third toss. Then
$A=\left\{\left(\omega_{1}, \omega_{2}, \omega_{3}, \ldots,\right)| \omega_{1}=T, \omega_{2}=T, \omega_{3}=H, \omega_{i} \in\{H, T\} \text { for } i>3\right\}.$


Given an event $A,$ let $A^{c}=\{\omega \in \Omega | \omega \notin A\}$ denote the **complement** of $A$.  Informally, $A^{c}$ can be read as "not $A$." The complement of $\Omega$ is the empty set $\emptyset$. The **union** of events $A$ and $B$ is defined as 
$$
A\cup B=\{\omega \in \Omega|\omega\in A\text{ or }\omega \in B\text{ or }\omega\in\text{ both}\}
$$ 
which can be thought of as "$A$ or $B$."  If $A_{1}, A_{2}, \ldots$ is a sequence of sets then
$$
\bigcup_{i=1}^{\infty} A_{i}=\left\{\omega \in \Omega | \omega \in A_{i} \text { for at least one i }\right\}.
$$
The **intersection** of $A$ and $B$ is defined as 
$$
A \cap B=\{\omega \in \Omega | \omega \in A\text{ and }\omega \in B\}
$$ 
which reads as "$A$ and $B$." Often $A \cap B$ is also written shortly as $AB$ or as $A,B.$  

If $A_{1}, A_{2}, \ldots$ is a sequence of sets then
$$
\bigcap_{i=1}^{\infty} A_{i}=\left\{\omega \in \Omega | \omega \in A_{i} \text { for all i }\right\}.
$$

If every element of $A$ is also contained in $B$ we write $A \subset B$ or, equivalently, $B \supset A$.  If $A$ is a finite set, let $|A|$ denote the number of elements in $A .$ We say that $A_{1}, A_{2}, \ldots$ are **disjoint** or **mutually exclusive** if $A_{i} \cap A_{j}=\emptyset$ whenever $i \neq j$. For example, $A_{1}=[0,1), A_{2}=[1,2), A_{3}=[2,3), \ldots$ are disjoint. A **partition** of $\Omega$ is a sequence of disjoint sets $A_{1}, A_{2}, \ldots$ such that $\bigcup_{i=1}^{\infty} A_{i}=\Omega$.

**Summary:** Sample space and events


$$
\begin{array}{ll}
\Omega & \text { sample space } \\
\omega & \text { outcome, elementary event, realization}\\
A      & \text { event (subset of } \Omega) \\
|A|    & \text { number of points in } A \text { (if } A \text { is finite) }\\
A^{c}  & \text { complement of } A (\operatorname{not} A)\\
A \cup B &\text{ union }(A\text{ or }B)\\
A \cap B &\text{ intersection }(A \text { and } B);\text{ short notations: }AB\text{ or }A,B\\
A \subset B &\text{ set inclusion }(A \text{ is a subset of or equal to }B)\\
\emptyset   &\text{ null event (always false)}\\
\Omega      &\text{ true event (always true)}
\end{array}
$$


  

<!-- Given an event $A,$ define the indicator function of $A$ by -->
<!-- $$ -->
<!-- I_{A}(\omega)=I(\omega \in A)=\left\{\begin{array}{ll} -->
<!-- 1 & \text { if } \omega \in A \\ -->
<!-- 0 & \text { if } \omega \notin A -->
<!-- \end{array}\right. -->
<!-- $$ -->
<!-- A sequence of sets $A_{1}, A_{2}, \ldots$ is monotone increasing if $A_{1} \subset A_{2} \subset$ -->
<!-- $\cdots$ and we define $\lim _{n \rightarrow \infty} A_{n}=\bigcup_{i=1}^{\infty} A_{i} .$ A sequence of sets $A_{1}, A_{2}, \ldots$ is -->
<!-- monotone decreasing if $A_{1} \supset A_{2} \supset \cdots$ and then we define $\lim _{n \rightarrow \infty} A_{n}=$ -->
<!-- $\bigcap_{i=1}^{\infty} A_{i} .$ In either case, we will write $A_{n} \rightarrow A$ -->

<!-- Example 2.4 Let $\Omega=\mathbb{R}$ and let $A_{i}=[0,1 / i)$ for $i=1,2, \ldots$ Then $\bigcup_{i=1}^{\infty} A_{i}=$ [0,1) and $\bigcap_{i=1}^{\infty} A_{i}=\{0\} .$ If instead we define $A_{i}=(0,1 / i)$ then $\bigcup_{i=1}^{\infty} A_{i}=$ -->
<!-- (0,1) and $\bigcap_{i=1}^{\infty} A_{i}=\emptyset$ -->

###  Probability

We want to assign a real number $P(A)$ to every event $A,$ called the **probability** of $A .$ We also call $P$ a **probability distribution** or a **probability measure**.  To qualify as a probability, $P$ has to satisfy three axioms. That is, a function $P$ that assigns a real number $P(A)\in[0,1]$ to each event $A$ is a **probability distribution** or a **probability measure** if it satisfies the following three axioms:

- **Axiom 1:** $P(A) \geq 0$ for every $A$ 
- **Axiom 2:** $P(\Omega)=1$
- **Axiom 3:** If $A_{1}, A_{2}, \ldots$ are disjoint then

$$
P\left(\bigcup_{i=1}^{\infty} A_{i}\right)=\sum_{i=1}^{\infty} P\left(A_{i}\right).
$$

::: {.callout-note}

## Sigma-Algebra 

It is not always possible to assign a probability to *every* event $A$ if the sample space is large. For instance, in the case of 
$$
\Omega=\mathbb{R}
$$ 
strange things can happen. There are pathological sets (e.g. Vitali sets) that simply break down the mathematics since they are non-measurable (i.e. we cannot assign probabilities to them).  Therefore, in cases like $\Omega=\mathbb{R}$, we assign probabilities to a *limited* class of sets called a **$\sigma$-field** or **$\sigma$-algebra**. 

For $\Omega=\mathbb{R}$, the canonical **$\sigma$-algebra** is the **Borel $\sigma$-algebra**. The Borel $\sigma$-algebra on $\mathbb{R}$ is generated by the collection of **all open subsets of $\mathbb{R}$**. 

In this course, we use the Borel $\sigma$-algebra and fortunately we do not have to bother with it any further. It's there lurking in the background making things work for us. 
::: 

One can derive many properties of $P$ from the three axioms. Here are a few: 

* $P(\emptyset)=0$
* $A \subset B\Rightarrow P(A) \leq P(B)$
* $0 \leq P(A) \leq 1$
* $P\left(A^{c}\right)=1-P(A)$
* $A \cap B=\emptyset \Rightarrow P(A \cup B)=P(A)+P(B)$

A less obvious property is given in the following: For any events $A$ and $B$ we have that, 

$$
P(A \cup B)=P(A)+P(B)-P(A B).
$$

**Example:** Two consecutive coin tosses. Let $H_{1}$ be the event that heads occurs on toss 1 and let $H_{2}$ be the event that heads occurs on toss 2. Likewise for $T_1$ and $T_2.$ Let us consider "fair" coins, i.e., all outcomes are equally likely. Then  
<!-- $$
P\left(\{H_{1},H_2\}\right)=P\left(H_{1}\right)+P\left(H_{2}\right)=\frac{1}{2}+\frac{1}{2}=1
$$
$$
P\left(\{T_{1},T_2\}\right)=P\left(T_{1}\right)+P\left(T_{2}\right)=\frac{1}{2}+\frac{1}{2}=1
$$ -->
<!-- $$
P\left(H_{1}\right)=P\left(H_{2}\right)=P\left(T_{1}\right)=P\left(T_{2}\right)=1 / 2
=P\left(\left\{H_{1}, T_{2}\right\}\right)+P\left(\left\{T_{1}, H_{2}\right\}\right)+P\left(\left\{T_{1}, T_{2}\right\}\right)
$$  -->
<!-- and -->
$$
\begin{align*}
%&P\left(\left\{H_{1}, H_{2}\right\} \cup
%        \left\{H_{1}, T_{2}\right\} \cup
%        \left\{T_{1}, H_{2}\right\} \cup
%        \left\{T_{1}, T_{2}\right\}\right)\\[2ex]
&P\left(\left\{H_{1}, H_{2}\right\}\right)+
  P\left(\left\{H_{1}, T_{2}\right\}\right)+
  P\left(\left\{T_{1}, H_{2}\right\}\right)+
  P\left(\left\{T_{1}, T_{2}\right\}\right)=\\[2ex]
&=\frac{1}{4}+\frac{1}{4}+\frac{1}{4}+\frac{1}{4}=1.               
\end{align*}
$$ 
Therefore,  
$$
\begin{align*}
P\left(H_{1} \cup H_{2}\right)&=P\left(H_{1}\right)+P\left(H_{2}\right)-P\left(H_{1} H_{2}\right)\\[2ex]
&=\frac{1}{2}+\frac{1}{2}-\frac{1}{4}=\frac{3}{4}.
\end{align*}
$$


#### Probabilities as relative frequencies {-}

One can interpret $P(A)$ in terms of **frequencies**. That is, $P(A)$ is the (infinitely) long run proportion of times that $A$ is true in repetitions. For example, if we say that the probability of heads is $1 / 2$, i.e $P(H)=1/2$ we mean that if we flip the coin many times then the proportion of times we get heads tends to $1 / 2$ as the number of tosses increases.  An infinitely long, unpredictable sequence of tosses whose limiting proportion tends to a constant is an idealization, much like the idea of a straight line in geometry. 


The following `R` codes approximates the probability $P(H)=1/2$ using 5, 50 and 5,000 many (pseudo) random coin flips:
```{r}
set.seed(869)
## 5 (fair) coin-flips:
results <- sample(x = c("H", "T"), size = 5, replace = TRUE)
## Relative frequency of "H" in 5 coin-flips
length(results[results=="H"])/5

## 50 (fair) coin-flips:
results <- sample(x = c("H", "T"), size = 50, replace = TRUE)
## Relative frequency of "H" in 50 coin-flips
length(results[results=="H"])/50

## 5000 (fair) coin-flips:
results <- sample(x = c("H", "T"), size = 5000, replace = TRUE)
## Relative frequency of "H" in 5000 coin-flips
length(results[results=="H"])/5000
```


### Independent Events

If we flip a fair coin twice, then the probability of two heads is $\frac{1}{2} \times \frac{1}{2}$.  We multiply the probabilities because we regard the two tosses as independent.  Two events $A$ and $B$ are called **independent** if 

$$
P(A B)=P(A) P(B).
$$

Or more generally, a whole set of events $\{A_i|i\in I\}$ is independent if 

$$
P\left(\bigcap_{i \in J} A_{i}\right)=\prod_{i \in J}P\left(A_{i}\right)
$$
for every finite subset $J$ of $I$, where $I$ denotes the not necessarily finite index set $($e.g. $I=\{1,2,\dots\}).$

 

Independence can arise in two distinct ways. Sometimes, we **explicitly assume** that two events are independent. For example, in tossing a coin twice, we usually assume the tosses are independent which reflects the fact that the coin has no memory of the first toss.

In other instances, we **derive** independence by verifying that the definition of independence $P(A B)=P(A)P(B)$ holds. For example, in tossing a fair die *once*, let $A=\{2,4,6\}$ be the event of observing an even number and let $B=\{1,2,3,4\}$ be the event of observing no $5$ and no $6$. Then, 
$A \cap B=\{2,4\}$ is the event of observing either a $2$ or a $4$. Are the events $A$ and $B$ independent?  
$$
P(A B)=\frac{2}{6}=P(A)P(B)=\frac{1}{2}\cdot \frac{2}{3}
$$ 
and so $A$ and $B$ are independent. In this case, we didn't assume that $A$ and $B$ are independent it just turned out that they were. 


**Cautionary Note:** Suppose that $A$ and $B$ are **disjoint events**, i.e. 
$$
AB=\emptyset,
$$ 
each with positive probability, i.e., $P(A)>0$ and $P(B)>0.$  Can they be independent?  No!  This follows since 

$$
P(A B)=P(\emptyset)=0\neq \underbrace{P(A)}_{>0}\;\underbrace{P(B)}_{>0}>0.
$$

Generally, there is no way to judge (in-)dependence by looking at the sets in a Venn diagram.


**Summary:** Independent Events


1. $A$ and $B$ are independent if $P(A B)=P(A) P(B)$.
2. Independence is sometimes assumed and sometimes derived.
3. Disjoint events with strictly positive probabilities are *not* independent.



### Conditional Probability

If $P(B)>0$ then the **conditional probability** of $A$ given $B$ is
$$
P(A \mid B)=\frac{P(A B)}{P(B)}.
$$
Think of $P(A \mid B)$ as the fraction of times $A$ occurs among those in which $B$ occurs. Here are some facts about conditional probabilities: 

* The rules of probability apply to events on the left of the bar "$\mid$". That is, for any fixed $B$ such that $P(B)>0,$ $P(\cdot \mid B)$ is a probability, i.e., it satisfies the three axioms of probability: 
   * $P(A \mid B) \geq 0$ 
   * $P(\Omega \mid B)=1$ 
   * If $A_{1}, A_{2}, \ldots$ are disjoint then $P\left(\bigcup_{i=1}^{\infty} A_{i} \mid B\right)=\sum_{i=1}^{\infty} P\left(A_{i} \mid B\right).$
* But it's generally **not true** that $P(A \mid B \cup C)=P(A \mid B)+P(A \mid C).$ 

In general it is also not the case that $P(A \mid B)=P(B \mid A)$. People get this confused all the time. For example, the probability of spots given you have measles is 1 but the probability that you have measles given that you have spots is not $1 .$ In this case, the difference between $P(A \mid B)$ and $P(B \mid A)$ is obvious but there are cases where it is less obvious. This mistake is made often enough in legal cases that it is sometimes called the **"prosecutor's fallacy"**.


**Example:** A medical test for a disease $D$ has outcomes $+$ and $-.$ The probabilities are:

$$
\begin{array}{c|cc|c} 
& D & D^{c} \\
\hline
+ & .0081 & .0900 &  .0981\\
- & .0009 & .9010 &  .9019\\
\hline
  & .0090 & .9910 &  1
\end{array}
$$
From the definition of conditional probability, we have: 

- Sensitivity of the test:
$$
\begin{align*}
P(+\mid D) 
& = \frac{P(+\cap D)}{P(D)}\\[2ex]
& = \frac{0.0081}{(0.0081+0.0009)}=0.9
\end{align*} 
- Specificity of the test: 
\begin{align*}
P(-\mid D^{c})
& = \frac{P(-\cap D^{c})}{P(D^{c})}\\[2ex]
& = \frac{0.9010}{(0.9010+0.0900)}\approx 0.9
\end{align*}
$$

Apparently, the test is fairly accurate: 

* Sick people yield a positive test result 90 percent of the time. 
* Healthy people yield a negative test result about 90 percent of the time. 

Now, suppose you go for a test and get a positive result. What is the probability you have the disease? Most people answer $0.90=90\%$. However, this conclusion is a fallacy ("prosecutor's fallacy") and not correct.

The correct answer is 
$$
\begin{align*}
P(D \mid+)
& = \frac{P(+\cap D)}{P(+)}\\[2ex]
& = \frac{0.0081}{(0.0081+0.0900)}=0.08 = 8\%.
\end{align*} 
$$

The lesson here is that you need to compute the answer numerically. Don't trust your intuition.


#### Independence via Conditional Probabilities {-}

If $A$ and $B$ are **independent events** then
$$
P(A \mid B)=\frac{P(A B)}{P(B)}=\frac{P(A) P(B)}{P(B)}=P(A)
$$
So another **interpretation of independence** is that knowing $B$ doesn't change the probability of $A$.


From the definition of conditional probability we can write 
$$
P(A B)=P(A \mid B) P(B)
$$ 
and also
$$
P(A B)=P(B \mid A) P(A).
$$ 
Often, these formulas give us a convenient way to compute $P(A B)$ when $A$ and $B$ are not independent. 



**Example:** Draw two cards from a deck, without replacement. Let $A$ be the event that the first draw is Ace of Clubs and let $B$ be the event that the second draw is Queen of Diamonds. Then 
\begin{align*}
P(A B) 
&=P(A) \;\;P(B \mid A)\\[2ex]
&=(1 / 52)\; (1 / 51)
\end{align*}



**Summary:** Conditional Probability
    
  1. If $P(B)>0$ then $P(A \mid B)=P(A B)/P(B)$
  2. $P(\cdot \mid B)$ satisfies the axioms of probability, for fixed $B$.  In general, $P(A \mid \cdot)$ does not satisfy the axioms of probability, for fixed $A$.
  3. In general, $P(A \mid B) \neq P(B \mid A)$.
  4. $A$ and $B$ are independent if and only if $P(A \mid B)=P(A)$.


## Random Variables

Statistics and econometrics are concerned with data. How do we link sample spaces, events and probabilities to data? The link is provided by the concept of a **random variable**. A real-valued **random variable** is a mapping $X: \Omega \rightarrow \mathbb{R}$ that assigns a real number $X(\omega)\in\mathbb{R}$ to each outcome $\omega$.


At a certain point in most statistics/econometrics courses, the sample space, $\Omega$, is rarely mentioned and we work directly with random variables.  But you should keep in mind that the sample space is really there, lurking in the background.


**Example:** Flip a coin ten times. Let $X(\omega)$ be the number of heads in the sequence $\omega.$ For example, if $\omega=\text{HHTHHTHHTT}$ then $X(\omega)=6$. 


**Example:** Let $\Omega=\left\{(x, y)|x^{2}+y^{2} \leq 1\right\}$ be the unit disc. Consider drawing a point "at random" from $\Omega$.  A typical outcome is then of the form $\omega=(x, y) .$ Some examples of random variables are $X(\omega)=x, Y(\omega)=y, Z(\omega)=x+y, W(\omega)=\sqrt{x^{2}+y^{2}}$. 


Given a real-valued random variable $X\in\mathbb{R}$ and a subset $A$ of the real line ($A\subset\mathbb{R}$), define $X^{-1}(A)=\{\omega \in \Omega|X(\omega) \in A\}$. This allows us to link the probabilities on the random variable $X$, i.e. the probabilities we are usually working with, to the underlying probabilities on the events, i.e. the probabilities lurking in the background. 


**Example:** Flip a coin twice and let $X$ be the number of heads. Then, $P_X(X=0)=P(\{T T\})=1 / 4$, $P_X(X=1)=P(\{H T, T H\})=1 / 2$ and $P_X(X=2)=P(\{H H\})=1 / 4$. Thus, the events and their associated probability distribution, $P$, and the random variable $X$ and its distribution, $P_X$, can be summarized as follows:



$\omega$     | $P(\{\omega\})$ | $X(\omega)$ 
-------------|-----------------|-------------
$T T$        | $1 / 4$         | 0 
$T H$        | $1 / 4$         | 1 
$H T$        | $1 / 4$         | 1 
$H H$        | $1 / 4$         | 2


$x$ | $P_X(X=x)$
----|------------
0   | $1 / 4$
1   | $1 / 2$
2   | $1 / 4$

<!-- Try generalizing this to $n$ flips. -->
Here, $P_{X}$ is not the same probability function as $P$, because $P$ maps from the sample space events, $\omega$, to $[0,1],$ while $P_X$ maps from the random-variable events, $X(\omega),$ to $[0,1].$  We will typically forget about the sample space $\Omega$ and just think of the random variable as an experiment with real-valued (possible multivariate) outcomes. We will therefore write $P\left(X=x_{k}\right)$ instead of $P_{X}\left(X=x_{k}\right)$ to simplify the notation.


### Univariate Distribution and Probability Functions  

#### Cumulative Distribution Function {#sec-Defcdf}

The **cumulative distribution function (cdf)**
$$F_{X}: \mathbb{R} \rightarrow [0,1]$$ 
of a real-valued random variable $X\in\mathbb{R}$ is defined by
$$
F_{X}(x)=\mathbb{P}(X \leq x).
$$


You might wonder why we bother to define the cdf. The reason is that it effectively contains all the information
about the random variable. Indeed, let $X\in\mathbb{R}$ have cdf $F$ and let $Y\in\mathbb{R}$ have cdf $G$.  If $F(x)=G(x)$ for all $x\in\mathbb{R}$ then $P(X \in A)=P(Y \in A)$ for all $A\subset\mathbb{R}$. In order to denote that two random variables, here $X$ and $Y$, have the same distribution, one can write shortly $X\overset{d}{=}Y$. 


<!-- 
**Caution:** Equality in distribution, $X\overset{d}{=}Y$, does generally **not** mean equality in realizations, that is $X\overset{d}{=}Y \not\Rightarrow X(\omega)=Y(\omega)$ for all $\omega\in\Omega$.  
-->


**The defining properties of a cdf:** A function $F$ mapping the real line to $[0,1]$, short $F:\mathbb{R}\to[0,1],$ is called a cdf for some probability measure $P$ if and only if it satisfies the following three properties:

  1. $F$ is non-decreasing i.e. $x_{1}<x_{2}$ implies that $F\left(x_{1}\right) \leq F\left(x_{2}\right)$.
  
  2. $F$ is normalized: $\lim_{x\rightarrow-\infty} F(x)=0$ and $\lim_{x \rightarrow \infty} F(x)=1$
  
  3. $F$ is right-continuous, i. e. $F(x)=F\left(x^{+}\right)$ for all $x$, where 
  $$
  F\left(x^{+}\right)=\lim_{y\to x, y>x} F(y).
  $$



Alternatively to cumulative distribution functions one can use **probability (mass) functions** in order to describe the probability law of **discrete** random variables and **density functions** in order to describe the probability law of **continuous** random variables. 


#### Probability Functions for Discrete Random Variables

A random variable $X$ is *discrete* if it takes only countably many values 
$$
X\in\{x_{1}, x_{2}, \ldots\}.
$$ 
For instance, $X\in\{1,2,3\}$ or $X\in\{2,4,6,\dots\}$ or $X\in\mathbb{Z}$ or $X\in\mathbb{Q}$. 

We define the **probability function** or **probability mass function (pmf)** for $X$ by
$$
f_{X}(x)=\mathbb{P}(X=x)\quad\text{for all}\quad x\in\{x_1,x_2,\dots\}
$$



#### Density Functions for Continuous Random Variables

A random variable $X$ is *continuous* if there exists a function $f_{X}$ such that 

- $f_{X}(x)\geq 0\quad$ for all $\quad x$
- $\int_{-\infty}^{\infty}f_{X}(x)dx=1 \quad$ and 
- $\mathbb{P}(a<X<b)=\int_{a}^{b} f_{X}(x) dx\quad$ for every $\quad a\leq b$.

The function $f_{X}$ is called the **probability density function (pdf)** or short **density function**. 


For density functions, we have that
$$
F_{X}(x)=\int_{-\infty}^{x} f_{X}(t) dt\quad\text{and}\quad f_{X}(x)=F_{X}^{\prime}(x)
$$
at all points $x$ at which $F_{X}$ is differentiable.


### Multivariate Distribution and Probability Functions 

A $d$-dimensional random vector is a column-vector $X=(X_1,\dots,X_d)^\prime$, where each element is a univariate random variable. 


#### Multidimensional Distribution Function 


The **multivariate distribution function** $F$ is given by
$$
F(a_1,\dots,a_d)=P(X_1\le a_1,\dots,X_d\le a_d).
$$

```{r, fig.align="center"}
## Install the package if not installed yet
# install.packages("mnormt")

library(mnormt)

x1     <- seq(-5, 5, 0.25) 
x2     <- seq(-5, 5, 0.25)
mu     <- c(0, 0)
sigma  <- matrix(c(2, -1, -1, 2), nrow = 2)
f      <- function(x_1, x_2) pmnorm(cbind(x_1, x_2), mu, sigma)
f_x1x2 <- outer(x1, x2, f)

persp(x = x1, y = x2, z = f_x1x2, theta = -30, phi = 25, 
      shade = 0.75, col = "blue", expand = 0.5, r = 2, 
      ltheta = 25, ticktype = "detailed", 
      xlab = "x\u2081", ylab = "x\u2082", zlab = "")
```



#### Multidimensional Probability Mass Function 

**Discrete random vectors:** $X$ takes only countably many (i.e. discrete) values 
$$
X\in\{x_1, x_2,\dots\} 
$$
with 
$$
x_i=\left(\begin{matrix}x_{i1}\\\vdots \\ x_{id}\end{matrix}\right)\in\mathbb{R}^d
$$
for $i=1,2,\dots$

**Multidimensional probability mass function** 

$$
f_{X}(x)=\mathbb{P}(X=x)\quad\text{for all}\quad x\in\{x_1,x_2,\dots\}
$$

<!-- 
$p(\mathbf{x}_i)=P(X=\mathbf{x}_i)$ for $i=1,2,\dots$. That is, 
\begin{align*}
P(X\in [a_1,b_1]\times\dots\times [a_d,b_d])=
\sum_{\mathbf{x}_i\in [a_1,b_1]\times\dots\times [a_d,b_d]}p(\mathbf{x}_i).
\end{align*} 
-->



#### Multidimensional Density Function 


**Continuous random vectors:** $X$ takes values in $\mathbb{R}^d$ 

$$
X=\left(\begin{matrix}X_{1}\\ X_{2}\\ \vdots \\ X_{d}\end{matrix}\right)\in\mathbb{R}^d
$$

and has a **multidimensional density function** $f(x_1,\dots,x_d)$. That is, 
$$
\begin{align*}
&P(X\in [a_1,b_1]\times\dots\times [a_d,b_d])=\\[2ex]
=&\int\limits_{a_d}^{b_d}\dots \int\limits _{a_1}^{b_1}f(x_1,\dots,x_d)dx_1\dots dx_d.
\end{align*}
$$

Properties of **multivariate density** functions: 

- $\displaystyle f(x_1,\dots,x_d)\geq 0$\par
- $\displaystyle \int_{-\infty}^{\infty}\dots \int_{-\infty}^{\infty}f(x_1,\dots,x_d)dx_1\dots dx_d=1$



```{r, fig.align="center"}
## Load the package
library(mnormt)

x1     <- seq(-5, 5, 0.25) 
x2     <- seq(-5, 5, 0.25)
mu     <- c(0, 0)
sigma  <- matrix(c(2, -1, -1, 2), nrow = 2)
f      <- function(x_1, x_2) dmnorm(cbind(x_1, x_2), mu, sigma)
f_x1x2 <- outer(x1, x2, f)

persp(x = x1, y = x2, z = f_x1x2, theta = -30, phi = 25, 
      shade = 0.75, col = "blue", expand = 0.5, r = 2, 
      ltheta = 25, ticktype = "detailed", 
      xlab = "x\u2081", ylab = "x\u2082", zlab = "")
```


In the following we focus only on continuous random vectors, but the discrete cases are treated analogously.


#### Marginal Distribution Functions and Marginal Density Functions 

Let

$$
X=\left(\begin{matrix}X_{1}\\ X_{2}\\ \vdots \\ X_{d}\end{matrix}\right)\in\mathbb{R}^d
$$

Each random element, $X_j$, with $j=1,\dots,d$, of the random vector $X$ has its own **marginal distribution** $F_j$. This is just the univariate distribution of $X_j$ when ignoring all other random variables in $X$. Formally we have:

- **Marginal distribution function:** $F_j(x)=P(X_j\leq x)$
- **Marginal density function:** $f_j$, for instance, for $j=1$:

$$
f_1({\color{blue}x_1})=\int_{-\infty}^{\infty}\dots \int_{-\infty}^{\infty}f({\color{blue}x_1},x_2\dots,
x_d)dx_2\dots  dx_d
$$


```{r, fig.align="center", echo=FALSE, fig.width=5, fig.height=5}
library("scales")
par(lwd=2,mgp=c(1,1,0), family="serif")
# Modified to extract diagonal.
bivariate.normal <- function(x, mu, Sigma) 
  exp(-.5 * diag(t(x-mu) %*% solve(Sigma) %*% (x-mu))) / sqrt(2 * pi * det(Sigma))

mu    <- c(0,0)
Sigma <- matrix(c(1,.8,.8,1), nrow=2)

x1 <- seq(-3, 3, length.out=50)
x2 <- seq(-3, 3, length.out=50)

# plot(1:10,axes=FALSE, frame.plot=TRUE, lwd=1)

# z can now be calculated much easier.
z     <- bivariate.normal(t(expand.grid(x1,x2)),mu,Sigma)
dim(z)<-c(length(x1),length(x2))
contour(x1, x2, z, col="#4545FF", drawlabels=FALSE, nlevels=5,
        xlab=expression(x[2]), ylab=expression(x[1]),  lwd=1,xlim=c(-3,4),ylim=range(x2),frame.plot=FALSE,axes=FALSE,xaxs = "i", yaxs = "i")
image(x1, x2, z, col=gray(seq(1,.2,len=10)), add=TRUE)
contour(x1, x2, z, col=alpha("darkblue", 0.5), drawlabels=FALSE, nlevels=5, add=TRUE)
#box(lwd=.5)
axis(1,at = -3:3, labels=FALSE,lwd.ticks=1)
axis(2,labels=FALSE,lwd.ticks=1)

# Marginal probability distribution: http://mpdc.mae.cornell.edu/Courses/MAE714/biv-normal.pdf
# Please check this, I'm not sure it is correct.
marginal.x1<-function(x)  exp((-(x-mu[1])^2)/2*(Sigma[1,2]^2)) / (Sigma[1,2]*sqrt(2*pi))
marginal.x2<-function(x)  exp((-(x-mu[1])^2)/2*(Sigma[2,1]^2)) / (Sigma[2,1]*sqrt(2*pi))

# Left side solid
x.s  <- seq(from=min(x1),to=max(x1),by=0.1)
vals<-marginal.x2(x.s)
lines(vals-abs(min(x1)),x.s,lty=1,lwd=1, col="darkred")

# Bottom side solid
vals<-marginal.x1(x.s)
lines(x.s,vals-abs(min(x2)),lty=1,lwd=1, col="darkorange")
legend("topleft", col=c("darkred","darkorange"), lty=1, legend = c(expression(f[1]),expression(f[2])), bty="n",
       title = "Marginal Densities")
legend("topright", col=c("darkblue"), lty=1, legend = expression(f(x[1],x[2])), bty="n",
       title = "Joint Density")
```



#### Conditional Density Function {#sec-condDistr}

Often, we are interested in the **conditional distribution** of $X_j$ given certain values of all other random variables 

$$
X_1=x_1,\ldots, X_{j-1}=x_{j-1}, X_{j+1}=x_{j+1},\ldots,X_d=x_d.
$$
That is, the distribution of $X_j$ when fixing the values of  
$$
X_1=x_1,\ldots, X_{j-1}=x_{j-1}, X_{j+1}=x_{j+1},\ldots, X_d=x_d.
$$  


An important tool is here the **conditional density** of, for instance, $X_1$ given $X_2=x_2,\ldots,X_d=x_d$:
$$
f(x_1\mid x_2,\ldots,x_d)=\frac{f(x_1,x_2,\ldots,x_d)}{f_{X_{2},\ldots,X_{d}}(x_2,\ldots,x_d)},
$$
where $f_{X_{2},\ldots,X_{d}}$ denotes the joint density of $X_2,\ldots,X_d$.


```{r, fig.align="center", echo=FALSE, fig.width=5, fig.height=5}
par(lwd=2,mgp=c(1,1,0), family="serif")
# Modified to extract diagonal.
bivariate.normal <- function(x, mu, Sigma) 
  exp(-.5 * diag(t(x-mu) %*% solve(Sigma) %*% (x-mu))) / sqrt(2 * pi * det(Sigma))

mu    <- c(0,0)
Sigma <- matrix(c(1,.8,.8,1), nrow=2)

x1 <- seq(-3, 3, length.out=50)
x2 <- seq(-3, 3, length.out=50)

# plot(1:10,axes=FALSE, frame.plot=TRUE, lwd=1)

# z can now be calculated much easier.
# z     <- bivariate.normal(t(expand.grid(x1,x2)),mu,Sigma)
# dim(z)<-c(length(x1),length(x2))
# contour(x1, x2, z, col="#4545FF", drawlabels=FALSE, nlevels=5,
#         xlab=expression(X[1]), ylab=expression(X[2]),  lwd=1,xlim=c(-3,4),ylim=range(x2),frame.plot=FALSE,axes=FALSE,xaxs = "i", yaxs = "i")


z     <- bivariate.normal(t(expand.grid(x1,x2)),mu,Sigma)
dim(z)<-c(length(x1),length(x2))
contour(x1, x2, z, col="#4545FF", drawlabels=FALSE, nlevels=5,
        xlab=expression(x[2]), ylab=expression(x[1]),  lwd=1,xlim=c(-3,4),ylim=range(x2),frame.plot=FALSE,axes=FALSE,xaxs = "i", yaxs = "i")
image(x1, x2, z, col=gray(seq(1,.2,len=10)), add=TRUE)
contour(x1, x2, z, col=alpha("darkblue", 0.5), drawlabels=FALSE, nlevels=5, add=TRUE)

#box(lwd=.5)
axis(1,labels=FALSE,lwd.ticks=1)
axis(2,labels=FALSE,lwd.ticks=1)
abline(v=.7, col=1, lwd=1.1, lty=2)
text(2, -2, labels=expression(x[2]==0.7))

# Dotted
f    <- function(x1,x2) bivariate.normal(t(cbind(x1,x2)),mu,Sigma)
x.s  <- seq(from=min(x1),to=max(x1),by=0.1)
vals <- f(x1=0.7,x2=x.s)
lines(vals-abs(min(x1)),x.s,lty=2,lwd=1.2, col="darkgreen")


marginal.x1<-function(x) exp((-(x-mu[1])^2)/2*(Sigma[1,2]^2)) / (Sigma[1,2]*sqrt(2*pi))
marginal.x2<-function(x) exp((-(x-mu[1])^2)/2*(Sigma[2,1]^2)) / (Sigma[2,1]*sqrt(2*pi))

# Left side solid
vals<-marginal.x2(x.s)
lines(vals-abs(min(x1)),x.s,lty=1,lwd=1, col=gray(.5, 0.5))
 
# Bottom side solid
vals<-marginal.x1(x.s)
lines(x.s,vals-abs(min(x2)),lty=1,lwd=1, col=gray(.5, 0.5))

# legend("topright", col=c("darkred","darkorange"), lty=1, legend = c(expression(f[1]),expression(f[2])), bty="n",
#        title = "Marginal Densities")
legend("topleft", col="darkgreen", lty=2.1, lwd=.98, 
legend = expression(f(x[1]~ "|" ~ X[2]==0.7)), bty="n",
       title = "Conditional Density")
```


### Means and Moments

#### Unconditional Means

The **unconditional mean** of $X_1$ is given by
$$
E(X_1)= \int x f_{X_1}(x)dx.
$$
The unconditional mean of a random vector $X=(X_1,\dots,X_d)'$ is given by the vector of element-wise means
$$
E(X)=(E(X_1),\dots,E(X_d))'.
$$



#### Means of Transformed Random Variables and Moments


The **mean of a transformed random variable** $r(X)$ is given by 
$$
E(r(X))=\int r(x) f_{X}(x)dx.
$$
Typical transformations are, for instance

- centering $r(x)=x-E(X)$,
- centering and scaling $r(x)=(x-E(X))/\sqrt{Var(X)}$, 
- or $r(x)=(x - E(X))^2$, 

where the latter transformation leads to the **second central moment**, i.e. the variance of $X$, $Var(X)=\int (x - E(X))^2 f_{X}(x)dx$.



* The $k$th, $k>0$, moment is given by 
$$
\mu_{k}=\mathrm{E}\left[X^{k}\right]=\int_{-\infty}^{+\infty}x^{k} f_X(x)d x.
$$

* The $k$th, $k>1$, central moment is given by 
$$
\mu^c_{k}=\mathrm{E}\left[(X-\mathrm{E}[X])^{k}\right]=\int_{-\infty}^{+\infty}(x-\mu)^{k} f_X(x)d x,
$$
where $\mu=E(X)$. 


**Note:** Moments determine the tail of a distribution (but not much else); see @LB2000. Roughly: The more moments a distribution has the faster converge the tails to zero. Distributions with compact supports (e.g. the uniform distribution $\mathcal{U}[a,b]$) have infinitely many moments. The Normal distribution has also infinitely many moments, even though this distribution has not a compact support; see @sec-NormalDistr. 






#### Conditional Means 

Of central importance in **regression analysis** is the **conditional mean**.  The conditional mean of $X_1$ for given values  $X_2=x_2,\ldots,X_d=x_d$:
$$
\begin{align*}
  m(x_2,\dots,x_d):&=E(X_1|X_2=x_2,\ldots,X_d=x_d)\\
                   &= \int x_1 f(x_1\mid x_2,\ldots,x_d)dx_1,
\end{align*}                  
$$
where $m(x_2,\dots,x_d)$ denotes the **regression function**. 



#### Law of Total Expectation 

As long as we do not fix the values of the conditioning variables, $X_2,\dots,X_d$, they are random variables.  Consequently, the conditional mean is generally itself a random variable 
$$
E(X_1|X_2,\ldots,X_d)=\int x_1 f(x_1\mid X_2,\ldots,X_d)dx_1 
$$
due to the randomness in $X_2,\ldots,X_d$. 


The **Law of Total Expectation** or **Law of Iterated Expectation** states that 
$$
E\big(E(X_1|X_2,\ldots,X_d)\big)=E(X_1).
$$ 

To show this result, we, first compute the conditional mean of $X_1$ given $X_2,\dots,X_d$, and then mean with respect to $X_2,\dots,X_d.$
<!-- Note that $f(x_1\mid X_2,\ldots,X_d)$ is just a transformation of the random variables $X_2,\dots,X_d$. So we can easily compute the unconditional mean $E(X_1)$ by taking the mean of $E(X_1|X_2,\ldots,X_d)$ as following,  -->
$$
\begin{align*}
&E\big({\color{RedViolet}E(X_1|X_2,\ldots,X_d)}\big)=\\
&=\int\dots\int\;{\color{RedViolet}\int x_1 f(x_1\mid x_2,\ldots,x_d)dx_1}\;f_{X_2,\dots,X_d}(x_2,\ldots,x_d)dx_2\dots dx_d\\
&=\int x_1 \left({\color{blue}\int\dots\int f(x_1,x_2,\ldots,x_d)dx_2\dots dx_d}\right)dx_1\\
&=\int x_1 {\color{blue}f_{X_1}(x_1)}dx_1\\
&=E(X_1).
\end{align*}
$$



### Independent Random Variables 

Random variables $X_1,\dots,X_d$ are mutually **independent** if for all $x=(x_1,\dots,x_d)^\prime$ it is true that
$$
\begin{align*}
  F(x_1,\dots,x_d)&=F_1(x_1)\cdot F_2(x_2)\cdot\ldots\cdot F_d(x_d)\\
  f(x_1,\dots,x_d)&=f_1(x_1)\cdot f_2(x_2)\cdot\ldots\cdot f_d(x_d)
\end{align*}
$$

The following holds true:

- Two real-valued random variables $X$ and $Y$ are independent from each other *if and only if* the marginal density of $X$ equals the conditional density of $X$ given $Y=y$ for all $y\in\mathbb{R}$,
$$
f_X(x)=f_{X|Y}(x\mid y)\quad \text{ for all } y\in\mathbb{R}.
$$
Of course, the same statement applies to the marginal density of $Y$ given $X=x$ for all $x\in\mathbb{R}$. That is, $X$ and $Y$ are two independent real-valued random variables *if and only if* 
$f_Y(y)=f_{Y|X}(y\mid x)$ for all $x\in\mathbb{R}.$
- If a real-valued random variable $X$ is independent from a real-valued random variable $Y$, then the conditional mean of $X$ given $Y=y$ equals the unconditional mean of $X$ for all $y\in\mathbb{R}$ 
$$
E(X\mid Y=y)=E(X)
$$
and likewise
$$
E(Y\mid X=x)=E(Y)
$$
for all $x\in\mathbb{R}$. 


**Note:** The properties that $E(X\mid Y=y)=E(X)$ for all $y\in\mathbb{R}$ or that $E(Y\mid X=x)=E(Y)$ for all $x\in\mathbb{R}$, do **not** imply that $Y$ and $X$ are independent. It only means that $Y$ has no effect on the mean of $X$, but it may have, for instance, an effect on the variance of $X$. 





### Random Samples 

Tradition dictates that the sample size is denoted by the natural number $n\in\{1,2,\dots\}$. A collection of random variables $(X_{1}, \ldots, X_{n})$ is called a **random sample** if its random variables are **i.i.d. (independent and identically distributed)**, i.e., if

1. $X_{1}, \ldots, X_{n}$ are all **independent** from each other and 
2. $X_{1}, \ldots, X_{n}$ have **identical** marginal distributions, i.e., $X_i\sim F_X$ for all $i=1,\dots,n$. 

In micro-econometrics, random samples are the default sampling scheme. That is, we consider the collected data as a realization of an underlying random sample. 


### Some Important Discrete Random Variables

<!-- ### The point mass distribution {-}  -->

<!-- $X$ has a point mass distribution at $a$, written $X \sim \delta_{a},$ if $P(X=a)=1$ in which case -->
<!-- $$ -->
<!-- F(x)=\left\{\begin{array}{ll} -->
<!-- 0 & x<a \\ -->
<!-- 1 & x \geq a -->
<!-- \end{array}\right. -->
<!-- $$ -->
<!-- The probability function is $f(x)=1$ for $x=a$ and $0$ otherwise. -->


#### The Discrete Uniform Distribution 

Let $k>1$ be a given integer. Suppose that $X$ has probability mass function given by
$$
f(x)=\left\{\begin{array}{ll}
1 / k & \text { for } x=1, \ldots, k \\
0 & \text { otherwise. }
\end{array}\right.
$$
We say that $X$ has a uniform distribution on $\{1, \ldots, k\}$.

```{r}
set.seed(51)
## Set the parameter k
k <- 10
## Draw one realization from the discrete uniform distribution
sample(x = 1:k, size = 1, replace = TRUE)
```


#### The Bernoulli Distribution 

Let $X$ represent a possibly unfair coin flip.  Then $P(X=1)=p$ and $P(X=0)=1-p$ for some $p \in[0,1]$.  We say that $X$ has a Bernoulli distribution written $X\sim\operatorname{Bernoulli }(p)$.  The probability function is $f(x)=p^{x}(1-p)^{1-x}$ for $x \in\{0,1\}$

```{r}
set.seed(51)
## Set the parameter p
p <- 0.25
## Draw n realization from the discrete uniform distribution
n <- 5
sample(x = c(0,1), size = n, prob = c(1-p, p), replace=TRUE)

## Alternatively:
## (Bernoulli(p) equals Binomial(1,p))
rbinom(n = n, size = 1, prob = p)
```


#### The Binomial Distribution

Suppose we have a coin which falls heads with probability $p$ for some $p\in[0,1]$. Flip the coin $n$ times and let $X$ be the number of heads (or successes). Assume that the tosses are independent. Let $f(x)=P(X=x)$ be the mass function. It can be shown that
$$
f(x)=\left\{
\begin{array}{ll}
\left(\begin{array}{l}
n \\
x
\end{array}\right) p^{x}(1-p)^{n-x} & \text { for } x=0, \ldots, n \\
0 & \text { otherwise. }
\end{array}\right.
$$
A random variable with this probability mas function is called a **binomial random variable** and we write $X \sim \operatorname{Binomial}(n, p)$.  

If 
$X_{1} \sim$ Binomial $\left(n_1, p\right)$ and 
$X_{2} \sim$ Binomial $\left(n_2, p\right)$ and if 
$X_1$ and $X_2$ 
are independent, then 
$X_{1}+X_{2} \sim \operatorname{Binomial}\left(n_1 + n_2, p\right).$


```{r}
set.seed(51)
## Set the parameters n and p
size <-   10 # number of trials
p    <- 0.25 # prob of success

## Draw n realization from the binomial distribution:
n <- 5
rbinom(n = n, size = size, prob = p)
```


### Some Important Continuous Random Variables

#### The Uniform Distribution 

$X$ has a $\operatorname{Uniform}(a, b)$ distribution, written $X\sim \mathcal{U}(a, b),$ if
$$
f(x)=\left\{\begin{array}{ll}
\frac{1}{b-a} & \text { for } x \in[a, b] \\
0 & \text { otherwise }
\end{array}\right.
$$
where $a<b$. The distribution function is
$$
F(x)=\left\{\begin{array}{ll}
0 & x<a \\
\frac{x-a}{b-a} & x \in[a, b] \\
1 & x>b
\end{array}\right.
$$

```{r}
## Drawing from the uniform distribution:
n <- 10
a <- 0
b <- 1
runif(n = n, min = a, max = b)
```


#### The Normal (or Gaussian) Distribution {#sec-NormalDistr}

$X$ has a Normal (or Gaussian) distribution with parameters $\mu$ and $\sigma,$ denoted by $X \sim \mathcal{N}\left(\mu, \sigma^{2}\right),$ if
$$
f(x)=\frac{1}{\sigma \sqrt{2 \pi}} \exp \left\{-\frac{1}{2 \sigma^{2}}(x-\mu)^{2}\right\}, \quad x \in \mathbb{R}
$$
where $\mu \in \mathbb{R}$ and $\sigma>0.$  Later we shall see that $\mu$ is the "center" (or mean of the distribution and $\sigma$ is the "spread" (or standard deviation) of the distribution. The Normal plays an important role in probability and statistics. Many phenomena in nature have approximately Normal distributions.  The **Central Limit Theorem** gives a special role to the Normal distribution by stating that the distribution of averages of random variables can be approximated by a Normal distribution.


We say that $X$ has a standard Normal distribution if $\mu=0$ and $\sigma=1$. Tradition dictates that a standard Normal random variable is denoted by $Z$. The pdf and cdf of a standard Normal are denoted by $\phi(z)$ and $\Phi(z)$. There is no closed-form expression for $\Phi$. Here are some useful facts:

(i) If $X \sim \mathcal{N}\left(\mu, \sigma^{2}\right)$ then $Z=(X-\mu) / \sigma \sim \mathcal{N}(0,1)$

(ii) If $Z \sim \mathcal{N}(0,1)$ then $X=\mu+\sigma Z \sim \mathcal{N}\left(\mu, \sigma^{2}\right)$

(iii) If $X_{i} \sim \mathcal{N}\left(\mu_{i}, \sigma_{i}^{2}\right), i=1, \ldots, n$ are independent then
$$
\sum_{i=1}^{n} X_{i} \sim \mathcal{N}\left(\sum_{i=1}^{n} \mu_{i}, \sum_{i=1}^{n} \sigma_{i}^{2}\right).
$$

The following `R`-codes plots the standard Normal density function:
```{r, echo = T, eval = T, message = F, warning = F, fig.align='center'} 
# draw a plot of the N(0,1) pdf
curve(dnorm(x),
      xlim = c(-3.5, 3.5),
      ylab = "Density", 
      main = "Standard Normal Density Function") 
```

This is how you can draw realizations from pseudo random Normal variables:
```{r}
## Drawing from the uniform distribution:
n     <- 12
mu    <- 0
sigma <- 1
rnorm(n = n, mean = mu, sd = sigma) 
```

An extension of the normal distribution in a univariate setting is the multivariate normal distribution. Let $X=(X_1,\dots,X_k)'$ be a $k$-dimensional normal variable, short $X\sim N_k(\mu,\Sigma)$ with mean vector $E(X)=\mu\in\mathbb{R}^k$ and covariance matrix $\operatorname{Cov}(X)=\Sigma$. The joint density function or **probability density function (pdf)** of the $k$-dimensional multivariate normal distribution is 
$$
f_{X}\left(x_{1}, \ldots, x_{k}\right)=\frac{\exp \left(-\frac{1}{2}(x-\mu)' \Sigma^{-1}(x-\mu)\right)}{\sqrt{(2 \pi)^{k}|\Sigma|}},
$$
where $|\Sigma|$ denotes the determinant of $\Sigma$. For $k=2$ we have the bivariate pdf of two random normal variables, $X$ and $Y$ say
$$
\begin{align*}
&g_{X,Y}(x,y) = \frac{1}{2\pi\sigma_X\sigma_Y\sqrt{1-\rho_{XY}^2}} \\ 
& \cdot \, \exp \left\{ \frac{1}{-2(1-\rho_{XY}^2)} \left[ \left( \frac{x-\mu_x}{\sigma_X} \right)^2 - 2\rho_{XY}\left( \frac{x-\mu_X}{\sigma_X} \right)\left( \frac{y-\mu_Y}{\sigma_Y} \right) + \left( \frac{y-\mu_Y}{\sigma_Y} \right)^2 \right]  \right\}.
\end{align*}
$$
Lets consider the special case where $X$ and $Y$ are independent standard normal random variables with densities $f_X(x)$ and $f_Y(y)$. We then have the parameters $\sigma_X = \sigma_Y = 1$, $\mu_X=\mu_Y=0$ (due to marginal standard normality) and correlation $\rho_{XY}=0$ (due to independence). The joint density of $X$ and $Y$ then becomes
$$
g_{X,Y}(x,y) = f_X(x) f_Y(y) = \frac{1}{2\pi} \cdot \exp \left\{ -\frac{1}{2}\left[x^2 + y^2\right]\right\}.
$$ 


#### The Chi-Squared Distribution {#sec-chisqdist}

The chi-squared distribution is another distribution relevant in econometrics. It is often needed when testing special types of hypotheses frequently encountered when dealing with regression models. 

The sum of $\nu=1,2,\dots$ squared *independent standard normal* distributed random variables, $Z_1,\dots,Z_\nu$ follows a chi-squared distribution with $\nu$ degrees of freedom (df):
$$
\begin{align*}
Z_1^2 + \dots + Z_\nu^2 = \sum_{j=1}^\nu Z_j^2 =:W \sim \chi^2_{\nu}. 
\end{align*}
$$
A $\chi^2$ distributed random variable $W$ with $\nu$ degrees of freedom has expectation $E(W)=\nu$, mode at $\nu-2$ for $\nu \geq 2$ and variance $2 \cdot \nu.$ 


Using the code below, we can display the pdf and the distribution function or **cumulated density function (cdf)** of a $\chi^2_3$ random variable in a single plot. This is achieved by setting the argument `add = TRUE"` in the second call of `"curve()"`. Further we adjust limits of both axes using `"xlim"` and `"ylim"` and choose different colors to make both functions better distinguishable. The plot is completed by adding a legend with help of `"legend()"`.


```{r, echo = T, eval = T, message = F, warning = F, fig.align='center'} 
# plot the pdf
curve(dchisq(x, df = 3), 
      xlim = c(0, 10), 
      ylim = c(0, 1), 
      col = "blue",
      ylab = "",
      main = "pdf and cdf of Chi-Squared Distribution (df = 3)")

# add the cdf to the plot
curve(pchisq(x, df = 3), 
      xlim = c(0, 10), 
      add = TRUE, 
      col = "red")

# add a legend to the plot
legend("topleft", 
       c("pdf", "cdf"), 
       col = c("blue", "red"), 
       lty = c(1, 1))
```


Since the outcomes of a $\chi^2_\nu$ distributed random variable are always positive, the support of the related pdf and cdf is $\mathbb{R}_{\geq 0}$.

As expectation and variance depend (solely!) on the degrees of freedom, the distribution's shape changes drastically if we vary the degrees of freedoms. 


```{r, echo = T, eval = T, message = F, warning = F, fig.align='center'} 
# plot the density for M=1
curve(dchisq(x, df = 1), 
      xlim = c(0, 15), 
      xlab = "x", 
      ylab = "Density", 
      main = "Chi-Square Distributed Random Variables")

# add densities for M=2,...,7 to the plot using a 'for()' loop 
for (M in 2:7) {
  curve(dchisq(x, df = M),
        xlim = c(0, 15), 
        add = T, 
        col = M)
}

# add a legend
legend("topright", 
       as.character(1:7), 
       col = 1:7 , 
       lty = 1, 
       title = "df")
```


Increasing the degrees of freedom shifts the distribution to the right (the mode becomes larger) and increases the dispersion (the distribution's variance grows).



#### The Student $t$ Distribution {#sec-tdist}

Let $Z$ be a standard normal random variable, $W$ a $\chi^2_\nu$ random variable and further assume that $Z$ and $W$ are independent. Then it holds that
$$
\frac{Z}{\sqrt{W/\nu}} =:X \sim t_\nu
$$
and $X$ follows a *Student $t$ distribution* (or simply $t$ distribution) with $\nu$ degrees of freedom.

The shape of a $t_\nu$ distribution depends on $\nu$. $t$ distributions are symmetric, bell-shaped and look similar to a normal distribution, especially when $\nu$ is large. This is not a coincidence: for a sufficiently large $\nu$, the $t_\nu$ distribution can be approximated by the standard normal distribution. This approximation works reasonably well for $\nu\geq 30$. 

A $t_\nu$ distributed random variable $X$ has an expectation if $\nu>1$ and it has a variance if $\nu>2$.
$$
\begin{align*}
  E(X) =& 0, \qquad \nu > 1 \\
  \text{Var}(X) =& \frac{\nu}{\nu-2}, \qquad \nu > 2
\end{align*}
$$

Let us plot some $t$ distributions with different degrees of freedoms $\nu$ and compare them to the standard normal distribution.

```{r, echo = T, eval = T, message = F, warning = F, fig.align='center'} 
# plot the standard normal density
curve(dnorm(x), 
      xlim = c(-4, 4), 
      xlab = "x", 
      lty = 2, 
      ylab = "Density", 
      main = "Densities of t Distributions")

# plot the t density for M=2
curve(dt(x, df = 2), 
      xlim = c(-4, 4), 
      col = 2, 
      add = T)

# plot the t density for M=4
curve(dt(x, df = 4), 
      xlim = c(-4, 4), 
      col = 3, 
      add = T)

# plot the t density for M=25
curve(dt(x, df = 25), 
      xlim = c(-4, 4), 
      col = 4, 
      add = T)

# add a legend
legend("topright", 
       c("N(0, 1)", "df=2", "df=4", "df=25"), 
       col = 1:4, 
       lty = c(2, 1, 1, 1))
```

The plot illustrates that as the degrees of freedom increase, the shape of the $t$ distribution comes closer to that of a standard normal bell curve. Already for $\nu=25$ we find little difference to the standard normal density. If $\nu$ is small, we find the distribution to have heavier tails than a standard normal.


#### Cauchy Distribution

The Cauchy distribution is a special
case of the $t$ distribution corresponding to $\nu=1$. The density is
$$
f(x)=\frac{1}{\pi\left(1+x^{2}\right)}.
$$
<!-- To see that this is indeed a density, let's do the integral: -->
<!-- $$ -->
<!-- \begin{aligned} -->
<!-- \int_{-\infty}^{\infty} f(x) d x &=\frac{1}{\pi} \int_{-\infty}^{\infty} \frac{d x}{1+x^{2}}=\frac{1}{\pi} \int_{-\infty}^{\infty} \frac{d \tan ^{-1}}{d x} \\ -->
<!-- &=\frac{1}{\pi}\left[\tan ^{-1}(\infty)-\tan ^{-1}(-\infty)\right]=\frac{1}{\pi}\left[\frac{\pi}{2}-\left(-\frac{\pi}{2}\right)\right]=1 -->
<!-- \end{aligned} -->
<!-- $$ -->

For the Cauchy distribution, the **expectation does not exist** -- that is, it has no mean.  Let's try to compute the mean of a Cauchy distribution and see what goes wrong. Its mean should be
$$
\mu=E(X)=\int_{-\infty}^{\infty} \frac{x d x}{\pi\left(1+x^{2}\right)}.
$$
In order for this improper integral to exist, we need both integrals $\int_{-\infty}^{0}$ and $\int_{0}^{\infty}$ to be finite. Let's look at the second integral.
$$
\int_{0}^{\infty} \frac{x d x}{\pi\left(1+x^{2}\right)}=\left.\frac{1}{2 \pi} \log \left(1+x^{2}\right)\right|_{0} ^{\infty}=\infty
$$
Similarly, the other integral, $\int_{-\infty}^{0},$ is $-\infty$. Since they're not both finite, the integral $\int_{-\infty}^{\infty}$ doesn't exist. In other words $\infty-\infty$ is not a number. Thus, the Cauchy distribution has no mean. 

What this means in practice is that if you take a sample $x_{1}, x_{2}, \ldots, x_{n}$ from the Cauchy distribution, then the average $\bar{x}$ does not tend to a particular number. Instead, every so often you will get such a huge number, either positive or negative, that the average is overwhelmed by it.



#### The F Distribution {#sec-Fdist}

Another ratio of random variables important to econometricians is the ratio of two independent $\chi^2$ distributed random variables that are divided by their degrees of freedom $m$ and $n$. The quantity 

$$
\frac{W/m}{V/n} \sim F_{m,n} \ \ \text{with} \ \ W \sim \chi^2_m \ \ , \ \ V \sim \chi^2_n 
$$
follows an $F$ distribution with numerator degrees of freedom $m$ and denominator degrees of freedom $n$, denoted $F_{m,n}$. The distribution was first derived by George Snedecor but was named in honor of [Sir Ronald Fisher](https://en.wikipedia.org/wiki/Ronald_Fisher).

By definition, the support of both pdf and cdf of an $F_{m,n}$ distributed random variable is $\mathbb{R}_{\geq0}$.

Say we have an $F$ distributed random variable $Y$ with numerator degrees of freedom $3$ and denominator degrees of freedom $14$ and are interested in $P(Y \geq 2)$. This can be computed with help of the function `"pf()"`. By setting the argument `"lower.tail"` to `"FALSE"` we ensure that `R` computes $1- P(Y \leq 2)$, i.e,the probability mass in the tail right of $2$. 

```{r, echo = T, eval = T, message = F, warning = F}
pf(2, df1 = 3, df2 = 14, lower.tail = F)
```

We can visualize this probability by drawing a line plot of the related density and adding a color shading with `"polygon()"`.

```{r, echo = T, eval = T, message = F, warning = F, fig.align='center'}
# define coordinate vectors for vertices of the polygon
x <- c(2, seq(2, 10, 0.01), 10)
y <- c(0, df(seq(2, 10, 0.01), 3, 14), 0)

# draw density of F_{3, 14}
curve(df(x ,3 ,14), 
      ylim = c(0, 0.8), 
      xlim = c(0, 10), 
      ylab = "Density",
      main = "Density Function")

# draw the polygon
polygon(x, y, col = "orange")
```

The $F$ distribution is related to many other distributions. An important special case encountered in econometrics arises if the denominator degrees of freedom are large such that the $F_{m,n}$ distribution can be approximated by the $F_{m,\infty}$ distribution which turns out to be simply the distribution of a $\chi^2_m$ random variable divided by its degrees of freedom $m,$ i.e. 
$$ 
\frac{W}{m} \sim F_{m,\infty} \quad\text{with}\quad W \sim \chi^2_m.
$$


## Exercises

* [Exercises for Chapter 2](https://www.dropbox.com/scl/fi/id6hc7q13veovkopf6xnt/Ch2_Exercises1.pdf?rlkey=fhyv5m2fjw6p9kv2strkxomun&st=jiltpae4&dl=0)

* [Exercises for Chapter 2 with Solutions](https://www.dropbox.com/scl/fi/q4x4q4q996mav47sax8nf/Ch2_Exercises_with_Solutions1.pdf?rlkey=69osulookyvx96ttvj4ovo66a&st=03oydfn1&dl=0)

## References {-}
