<!-- LTeX: language=en-US -->
```{r setup, include=FALSE, cache=FALSE}
source("_common.R")
```

# Organization of the Course {-}

### Timetable {-}

```{r, echo=FALSE}
suppressPackageStartupMessages(library("kableExtra"))
suppressPackageStartupMessages(library("tidyverse"))

timetable <- tibble("Day"          = c("TBA",       "TBA"),
                    "Time (UTC+2)" = c("TBA",       "TBA"),
                    "Lecture Hall" = c("TBA",       "TBA"))

timetable %>% kbl() %>%  kable_styling()
```

::: {.callout-warning title="TODO — KU"}
Fill in the KU lecture days, times and rooms above. The Bonn slots were
Monday 10:15--11:45 and Wednesday 14:15--15:45 in Jur / Lecture Hall F.
:::

<!-- TODO — KU: Bonn streamed the first weeks via a uni-bonn Zoom room. Decide
     whether KU streams at all; if so, restore this section with a KU link.

#### Zoom

For students not yet on campus, I'll stream the lectures via [Zoom](ADD-KU-ZOOM-LINK) during the first lecture weeks.
-->

### Lecture Material  

* This online script

::: {.callout-warning title="TODO — KU"}
The eWhiteboard and `R`-scripts links pointed at the Bonn WS 25/26 Dropbox
folder (`Metrics_eWhiteboard_2526_annotated.pdf`) and are removed. Add the KU
equivalents once the semester folder exists:

* `[eWhiteboard](ADD-KU-LINK)`
* ``[Folder containing `R`-Scripts](ADD-KU-LINK)``
:::

### Literature {-}

The script is self-contained. To prepare well for the exam, it's a good idea to work with this script. 

Further textbooks I can recommend:  

* *All of Statistics: A Concise Course in Statistical Inference*, by Larry Wasserman
* *Econometrics*, by F. Hayashi
* *Econometrics*, by Bruce E. Hansen
* *Probability and Statistics for Economists*, by Bruce E. Hansen
* *Econometric theory and methods*, by R. Davidson and J.G. MacKinnon
<!-- * *Introduction to econometrics*, by J. Stock and M.W. Watson
  * TODO — KU: the e-book link here was a Bonn BONNUS/ULB catalogue record.
    Replace with the KU library equivalent if this entry is restored. -->
<!-- * *A primer in econometric theory*, by J. Stachurski -->


### Information on the Exam {-}

::: {.callout-warning title="TODO — KU"}
Everything in this section is the **Bonn** exam format and has not been
confirmed for KU. Check each item against the KU module description before the
semester starts, and replace the examinations-office link below.
:::

* Written exam (60 minutes):
   * **10** multiple-choice problems and
   * **2** free-text problems (each with sub-questions)
* In the exam you are allowed to use
   * a *handwritten* cheat-sheet (**one side** of a DIN A4 page).
   * a non-programmable scientific calculator.
* Please, do **not** use a pencil.
* Exam dates and times are published by the examinations office (TODO — KU: add link)
<!-- * There will be a **Mock Exam** in the last lecture weeks -->

<!-- TODO — KU: the mock exam below linked to the Bonn WS 25/26 Dropbox
     (Mock_Exam1.pdf). Restore with a KU mock exam when one exists.

* **Mock Exam** with 5 multiple-choice problems and 1 free-text problem: [Download](ADD-KU-LINK)
   * Note: The actual exam will contain **10** multiple-choice problems and **2** free-text problems.
-->



